Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-19 treestats_paper.tex
Give example of a "non-AFS" summary to give reader idea.

A key area of inefficiency in existing methods is that they focus on representing
the genotype data as a \emph{matrix}.  As such, computational complexity if often linear
in the matrix size. An alternative would be to take advantage of the relatedness amongst
individuals in the sample and represent the data as trees.
Trees describe the ancestral relationships between
samples~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encode genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.

>>>>

Fortunately, the genotype matrix is not the only way to encode genetic data.
Trees describe the ancestral relationships between
samples~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encode genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
